\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{hyperref}

\title{Hardware-Aware NPU Scheduling with Adaptive Optimization and Quantum-Assisted Refinement}
\author{Student Research Project}
\date{}

\begin{document}
\maketitle

\section{Introduction}
The project started from a practical issue we kept seeing in NPU scheduling: a schedule that looks fine at graph level can still perform poorly once SRAM pressure, bandwidth bursts, and queue backlogs are accounted for. A purely local heuristic may serialize work too early, evict data that should have stayed on-chip, or create memory traffic spikes that later show up as stalls.

So instead of building a single ``smart'' scheduler, we built a layered system. First, multiple classical strategies produce candidate topological orders. Then a hardware-aware cost model scores them. Finally, an adaptive penalty mechanism (APR) reshapes constraint importance over rounds, and a quantum-style refinement stage searches harder neighborhoods from a warm start.

\section{What Is Actually Implemented}
The repository is modular, and each file has a clear role:
\begin{itemize}
    \item \texttt{memory\_hierarchy.py}: SRAM residency simulation with eviction policy, prefetch credit, spill tracking, bank conflict cycles, and capacity violations.
    \item \texttt{bandwidth\_estimator.py}: read/write channel utilization, burst-window averaging, backlog carry-over, congestion and pipeline stall estimation.
    \item \texttt{fusion\_logic.py}: rule-based pair/triplet fusion gain estimation (e.g., conv--batchnorm--relu style chains).
    \item \texttt{cost\_model.py}: combines memory, bandwidth, fusion, and parallelism terms into total cost plus penalty terms.
    \item \texttt{penalty\_tuner.py}: APR update loop using violation rate, cost impact, and recurrence frequency.
    \item \texttt{scheduling\_engine.py}: Greedy, Lookahead, Beam Search, and Simulated Annealing order generation.
    \item \texttt{run\_experiment.py}: end-to-end orchestration, trial selection, APR warm-up, quantum runs, result export.
\end{itemize}

\section{Cost Model (Code-Aligned)}
The objective in \texttt{cost\_model.py} is not a single scalar heuristic; it is a structured sum of physically motivated terms:
\[
\begin{aligned}
C_{\text{raw}} =
&\;w_d\big(\text{dram\_access} + 0.08\cdot\text{spill\_count}\cdot\max(1,\text{avg\_sram\_util})\big) \\
&+ w_r\big(\text{sram\_reuse\_loss} + 0.45\cdot\text{bank\_conflict\_cycles}\big) \\
&+ w_b\big(\text{bandwidth\_congestion} + 0.35\cdot\text{backlog\_pressure}\big) \\
&+ w_s\big(\text{pipeline\_stalls} + 0.22\cdot\text{bank\_conflict\_cycles}\big) \\
&- w_f\cdot\text{fusion\_gain} + w_p\cdot\text{parallelism\_loss}
\end{aligned}
\]

Then penalties are added:
\[
\begin{aligned}
C_{\text{penalty}} =
&\;140\,M_{\text{sram}}v_{\text{sram}} + 100\,M_{\text{bw}}v_{\text{bw}} + 900\,M_{\text{dep}}v_{\text{dep}} \\
&+ 70\,M_{\text{dram}}\max(0, v_{\text{dram}}-1)
+ 90\,M_{\text{bank}}v_{\text{bank}}
+ 65\,M_{\text{imb}}v_{\text{imb}}
\end{aligned}
\]
\[
C_{\text{total}} = C_{\text{raw}} + C_{\text{penalty}}
\]

\textbf{Why these terms were chosen:}
\begin{itemize}
    \item \textbf{DRAM term}: captures expensive off-chip movement and spill side effects.
    \item \textbf{SRAM reuse loss}: captures re-fetch pain caused by eviction or missed reuse windows.
    \item \textbf{Bandwidth congestion + stalls}: models burst pressure and how it directly delays execution.
    \item \textbf{Fusion gain}: rewards schedules that place fusible operator chains adjacently.
    \item \textbf{Parallelism loss}: discourages orders that serialize the frontier even when parallel-ready work exists.
\end{itemize}

In this implementation, \texttt{parallelism\_loss} is critical-path-aware: when many nodes are ready but one branch is chosen repeatedly, the model adds serialization and slack penalties so the scheduler does not accidentally collapse available concurrency.

\section{APR in This Project}
APR is implemented in \texttt{penalty\_tuner.py}. Each round updates penalty multipliers from three signals:
\begin{itemize}
    \item \texttt{violation\_rate}: how much a constraint is being violated right now.
    \item \texttt{cost\_impact}: how strongly that constraint's region contributes to total cost.
    \item \texttt{constraint\_frequency}: how often the same constraint keeps appearing across rounds.
\end{itemize}

The update is:
\[
\text{growth} = \exp(\text{violation\_rate} + \text{cost\_impact} + \text{constraint\_frequency})
\]
with practical guards:
\begin{itemize}
    \item exponential growth is clipped by \texttt{max\_growth = 2.4}
    \item momentum blending uses \texttt{0.18} to reduce oscillation
    \item penalties are clipped to \([0.2, 14.0]\)
    \item both violation and cost signals are EMA-smoothed (decay \texttt{0.72})
\end{itemize}

\textbf{Why this matters:} violation alone is too reactive, and cost impact alone can ignore rare but dangerous constraints. Frequency keeps the system from forgetting recurring bottlenecks.

\section{Memory and Bandwidth Modeling Choices}
\subsection*{Memory hierarchy}
The SRAM simulator keeps an ordered tensor cache and evicts using a score that balances remaining consumers, tensor size, and age. It also tracks prefetch credits, bank conflicts, peak SRAM usage, spill count, and idle-cycle impact from evictions.

This design gives the scheduler feedback on \emph{why} a schedule is bad: not just ``DRAM is high,'' but whether the issue comes from poor reuse timing, bank pressure, or outright capacity overrun.

\subsection*{Bandwidth estimator}
Bandwidth uses separate read/write capacities with burst windows and decayed backlog. Overflow contributes to both congestion and pipeline stalls, and read-write asymmetry amplifies penalties.

Without this, the optimizer can choose orders that look cheap in memory terms but create traffic waves that stall compute later.

\section{Real Experiment Flow (\texttt{run\_experiment.py})}
The script follows this pipeline:
\begin{enumerate}
    \item Load \texttt{config.yaml} and \texttt{example\_workload.json} (31-node DAG).
    \item Instantiate memory, bandwidth, fusion, and cost-model modules.
    \item Run Greedy, Lookahead, Beam Search, and Simulated Annealing. For each strategy, run multiple seeded trials and pick the best via
    \[
    \text{rank\_score} = \text{total\_cost} - 120 \cdot \text{feasibility}.
    \]
    \item Run Quantum (QAOA-style) refinement starting from the annealing schedule.
    \item Warm APR using strong classical candidates (lookahead/beam/anneal and APR-warm SA), then choose an APR seed order by feasibility-first ranking.
    \item Run Quantum+APR for multiple rounds, update penalties each round, and select the best trial.
    \item Save \texttt{outputs/schedules.json}, \texttt{outputs/metrics.txt}, and \texttt{outputs/explanations.txt}.
\end{enumerate}

\section{Main Results and Interpretation}
Latest run from the current code (\texttt{python run\_experiment.py}):

\begin{table}[h]
\centering
\begin{tabular}{lrrrrrr}
\toprule
Strategy & Cost & Latency & DRAM & BW(\%) & Idle & Feasible(\%) \\
\midrule
Greedy & 5669.65 & 3456.25 & 2226.24 & 86.74 & 353.10 & 51.61 \\
Lookahead & 4168.69 & 3390.81 & 1649.04 & 88.09 & 287.72 & 54.83 \\
Beam Search & 4169.26 & 3435.52 & 1383.84 & 87.71 & 332.39 & 51.61 \\
Annealing & 4443.10 & 3488.46 & 1383.84 & 87.05 & 385.31 & 51.61 \\
Quantum & 4439.67 & 3443.75 & 1649.04 & 87.43 & 340.64 & 54.83 \\
Quantum + APR & 4943.55 & 3390.81 & 1649.04 & 88.09 & 287.72 & 54.83 \\
\bottomrule
\end{tabular}
\caption{Comparison from current repository run.}
\end{table}

\textbf{Interpretation instead of just ranking:}
\begin{itemize}
    \item Lookahead is the best total-cost strategy in this run and ties best latency.
    \item Beam reduces DRAM compared to Lookahead, but its congestion/stall burden is higher, so total cost ends up nearly the same.
    \item Quantum baseline improves over Greedy but does not beat the strongest classical candidate on this workload.
    \item Quantum+APR improves constraint behavior vs Quantum baseline (lower idle and lower bandwidth+stall burden), but total cost rises because APR pushes several penalties to the clip ceiling, so \texttt{penalty\_cost} dominates.
\end{itemize}

\section{Ablation Study}
I ran targeted ablations to understand what each subsystem contributes.

\begin{table}[h]
\centering
\begin{tabular}{lrrr}
\toprule
Ablation Setting & Cost & Latency & Feasible(\%) \\
\midrule
Baseline (Lookahead, full model) & 4168.69 & 3390.81 & 54.83 \\
Remove APR (use Quantum without APR) & 4439.67 & 3443.75 & 54.83 \\
Remove fusion gain (same Lookahead order, fusion off) & 4414.48 & 3390.81 & 54.83 \\
Remove bandwidth model from search objective$^\dagger$ & 4469.35 & 3496.34 & 51.61 \\
\bottomrule
\end{tabular}
\caption{Ablation effects measured on current code.}
\end{table}

$^\dagger$ For this ablation, scheduling was done with bandwidth terms/penalties disabled, then evaluated using the full baseline model.

\textbf{What this tells us:}
\begin{itemize}
    \item \textbf{Without APR:} total cost is lower than Quantum+APR in this run, but APR still improves latency and traffic smoothness. Here APR is conservative and over-penalizes.
    \item \textbf{Without fusion:} cost increases by 245.79 (about 5.9\%), showing fusion offsets a meaningful portion of memory+compute overhead.
    \item \textbf{Without bandwidth modeling:} cost increases by about 7.2\%, latency by about 3.1\%, and feasibility drops 3.22 points. So bandwidth awareness is not optional.
\end{itemize}

\section{Design Insights}
\begin{itemize}
    \item Using multiple classical search styles is better than trusting one heuristic. Different methods fail in different ways.
    \item Separating memory and bandwidth models gives actionable diagnostics; combined black-box penalties were much harder to tune.
    \item The feasibility-aware trial rank (cost minus feasibility bonus) is simple but useful for choosing stable candidates.
    \item APR needs clipping/momentum. Without these guards, penalty multipliers can oscillate or explode.
    \item Quantum-style refinement works best as a local improver on strong warm starts, not as a standalone scheduler.
\end{itemize}

\section{Limitations}
\begin{itemize}
    \item Small-scale simulation only (single synthetic 31-node workload in current setup).
    \item No real NPU hardware-in-the-loop validation.
    \item Memory and bandwidth models are detailed but still approximate, not cycle-accurate microarchitecture simulation.
    \item Quantum stage is simulator-style, so claims are about search behavior, not hardware quantum speedup.
\end{itemize}

\section{Conclusion}
This project is strongest as a systems-oriented scheduling study: a real pipeline, explicit hardware signals, and measurable trade-offs. The main takeaway from the current code is that hardware-aware modeling and search diversity provide consistent gains, while APR gives useful constraint control but still needs better calibration to avoid over-penalization on this workload.

\end{document}
